---
title: "Task 1 Test"
output: html_document
date: "2024-11-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(warn=-1)
```

```{r parameters}
seed = 445
csv_file_one = "data1.csv"
packages = list("caret", 
                "randomForest")

```

```{r, include=FALSE}
set.seed(seed)

for (package in packages){
  if(!require(package, character.only = TRUE)){
    install.packages(package, character.only = TRUE)
  } else {
    library(package, character.only = TRUE)
  }
}
```


# Task 1: binary classification

```{r}
task_one_df <- read.csv(csv_file_one)

str(task_one_df)
```


## T1.1: exploratory data analysis and summary statistics (Peter?)

```{r}

```


## T1.2: training and evaluating on various classifiers

### Without PCA

#### Linear Discriminant Analysis (LDA) (Imar)

```{r}

```

#### Logistic classifier (Imar)

```{r}

```

#### Quadratic Discriminant Analysis (QDA) (Chi)

```{r}

```

#### Nearest Neighbor Classifier (k-NN) (Chi)

```{r}

```

#### Gradient Boosting Decision Trees (GBDT) (Jonathan)

```{r}

```

#### Random Forest (Jonathan)

```{r}

```

#### Support Vector Machine (SVM) (Peter)

```{r}

```


### With PCA

#### Linear Discriminant Analysis (LDA) (Imar)

```{r}

```

#### Logistic classifier (Imar)

```{r}

```

#### Quadratic Discriminant Analysis (QDA) (Chi)

```{r}

```

#### Nearest Neighbor Classifier (k-NN) (Chi)

```{r}

```

#### Gradient Boosting Decision Trees (GBDT) (Jonathan)

```{r}

```

#### Random Forest (Jonathan)

```{r}

```

#### Support Vector Machine (SVM) (Peter)

```{r}

```


## T1.3:

## T1.4:


# Task 2: feature selection

## T2.1:

## T2.2:

## T2.3:

