---
title: "Task 2 RmD"
output: html_document
date: "2024-11-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(warn=-1)
```

```{r parameters}
seed = 443
csv_file_two = "data2.csv.gz"
packages = list("caret", 
                "randomForest", 
                "tree"
                )

```

```{r, include=FALSE}
set.seed(seed)

for (package in packages){
  if(!require(package, character.only = TRUE)){
    install.packages(package, character.only = TRUE)
  } else {
    library(package, character.only = TRUE)
  }
}
```


# Task 2: feature selection

```{r}
task_two_df <- read.csv(csv_file_two)

str(task_two_df)
```

```{r}
table(task_two_df$label)
```


## T2.1: exploratory data analysis and summary statistics

```{r}

```


## T2.2: train and evaluate feature selection methods

Train and evaluate feature selection methods with the goal of achieving high 
balanced accuracy using a minimal number of selected features. Use three 
different approaches of your choice that were covered in the course.

```{r}

```


## T2.3: additional feature selection methods

Train and evaluate up to three additional feature selection methods, 
other than those in T2.2, which may include methods not covered in the course, 
with the goal of achieving high balanced accuracy using a minimal number of 
selected features. Provide an explanation of each new method.

For T2.2 and T2.3, your evaluation results should include the balanced accuracy 
and the number of selected features achieved by each method for various numbers 
of selected features. You may also compare the balanced accuracy achieved 
by different methods for a similar number of selected features.

```{r}

```

