---
title: "Task 2 RmD"
output: html_document
date: "2024-11-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(warn=-1)
```

```{r parameters}
seed = 443
csv_file_two = "data2.csv.gz"
training_size = 0.8 # Proportion of dataset that is to be split for training
packages = list("caret", 
                "randomForest", 
                "tree",
                "glmnet",
                "leaps"
                )

```

```{r, include=FALSE}
for (package in packages){
  if(!require(package, character.only = TRUE)){
    install.packages(package, character.only = TRUE)
  } else {
    library(package, character.only = TRUE)
  }
}
```


# Task 2: feature selection

```{r load_data}
set.seed(seed)

# load_data_start_time <- Sys.time()

task_two_df <- read.csv(csv_file_two)

# load_data_end_time <- Sys.time()
# 
# load_data_runtime <- load_data_end_time - load_data_start_time
# print(load_data_runtime) # 1.3 mins


# check_na_start_time <- Sys.time()
# 
# # Checking if there are any rows that have missing values... Keep this hashed unless you want to check, because it takes a while to run
# for (i in 1:nrow(task_two_df)) {
#   if (any(is.na(task_two_df[i, ]))) {
#     print(paste("Row", i, "contains NA values."))
#   }
# }
# 
# check_na_end_time <- Sys.time() # 16.8 mins

# check_na_runtime <- check_na_end_time - check_na_start_time
# print(check_na_runtime) 


# task_two_df$label_as_factor <- as.factor(task_two_df$label)

```


```{r EDA}

nrow(task_two_df) # Our data originally has 800 observations (compounds)
ncol(task_two_df) # Our data originally has 100,001 features (1 label as well as 50000 real variables and 50000 random probes, randomly permuted)

# Construct dataset of solely the binary features
task_two_df_no_label <- task_two_df[, -1]

# For each feature, determine the number of observations with "1s" and the variance of the feature
totals_vector <- sapply(task_two_df_no_label, sum)
variance_vector <- sapply(task_two_df_no_label, var)

quantile(totals_vector)
quantile(variance_vector)

# Histogram of the totals and the variance
hist(totals_vector, breaks = seq(min(totals_vector), max(totals_vector), by = 1),
     main = "Histogram of number of compounds per feature")
hist(variance_vector, breaks = seq(min(variance_vector), max(variance_vector), length.out = 1000),
     main = "Histogram of variance of compounds per feature")

### We notice that there are numerous features that are present in almost no observations
### As such, we investigate the possiblity of implementing a variance threshold to 
### restrict the number of features we retain
sum(totals_vector == 0)

# If we implement a variance threshold of 0.01 (corresponding to a feature only being present
# in a maximum of 8 out of 800 observations), we would remove 69,879 features
above_var_threshold <- totals_vector[variance_vector > .01]
min(above_var_threshold)
sum(variance_vector <= .01)

```


```{r drop_columns less than 0.01 variance}
# Remove features that have less than or equal to 0.01 variance
cols_below_var_threshold <- apply(task_two_df, 2, var) <= 0.01
table(cols_below_var_threshold)
task_two_df <- task_two_df[, !cols_below_var_threshold]

# After removing features with low variance, our data prior to feature selection 
# has 800 observations and now only 30,122 features
nrow(task_two_df)
ncol(task_two_df)

# Labels represent whether given compound binds to target site on thrombin 
# Possible values are -1 (compound did not bind) and 1 (compound did bind)
table(task_two_df$label) # 772 compounds did not bind and 78 compounds did bind

```


```{r train_test_split}
# Split Data into Test and Train (We should all use the same data)
train_split = as.integer(nrow(task_two_df)*training_size)

train_indices <- sample(1:nrow(task_two_df), train_split)
train_data <- task_two_df[train_indices,]
test_data <- task_two_df[-train_indices,]
```


## T2.1: exploratory data analysis and summary statistics (??? -- need to expand)


Our partial EDA is in above chunk of code where we plot histograms of the totals
and the variance amongst each of the 100,000 original features. Should this
still be expanded?


## T2.2: train and evaluate feature selection methods

Train and evaluate feature selection methods with the goal of achieving high 
balanced accuracy using a minimal number of selected features. Use three 
different approaches of your choice that were covered in the course.

### Lasso (Jonathan)

```{r lasso with balanced accuracy}
# Prepare the data
test_data_clean_matrix <- as.matrix(test_data[, -1])
train_data_matrix <- as.matrix(train_data[, -1])
train_labels <- train_data$label
test_labels <- test_data$label

# Define the lambda grid
grid <- 10^seq(-4, 4, length = 100)

# Function to calculate balanced accuracy
calculate_balanced_accuracy <- function(confusion) {
  if (all(dim(confusion) == c(2, 2))) {
    balanced_acc <- (1/2) * (confusion[2,2] / (confusion[2,2] + confusion[1,2])) + 
                    (1/2) * (confusion[1,1] / (confusion[1,1] + confusion[2,1]))
  } else {
    balanced_acc <- NA
  }
  return(balanced_acc)
}

# Creating a loop that will test the balanced accuracies for each of the different values of lambda taken from the grid above. 
start_task_two_lasso <- Sys.time()
results <- sapply(grid, function(lambda) {
  fit.lasso <- glmnet(train_data_matrix, train_labels, alpha = 1, lambda = lambda, family = "binomial")
  pred.test <- predict(fit.lasso, newx = test_data_clean_matrix, type ="class")
  # pred.test_binary <- ifelse(pred.test > 0, 1, -1)
  confusion <- table(Predicted = pred.test, Actual = test_labels)
  balanced_acc <- calculate_balanced_accuracy(confusion)
  num_predictors <- sum(coef(fit.lasso) != 0) - 1
  return(c(balanced_acc, num_predictors))
})
end_task_two_lasso <- Sys.time()
total_lasso_time <- end_task_two_lasso - start_task_two_lasso
print(total_lasso_time)

balanced_accuracies <- results[1, ]
num_predictors <- results[2, ]

# Plot balanced accuracies against lambda values
plot(grid, balanced_accuracies, type = "b", log = "x", xlab = "Lambda", ylab = "Balanced Accuracy", main = "Balanced Accuracy for Different Lambda Values")
grid()

# Plot number of predictors against lambda values
plot(grid, num_predictors, type = "b", log = "x", xlab = "Lambda", ylab = "Number of Predictors", main = "Number of Predictors for Different Lambda Values")
grid()

# Find the lambda with the highest balanced accuracy
best_lambda_index <- which(balanced_accuracies == max(balanced_accuracies, na.rm = TRUE))
best_lambda <- grid[best_lambda_index]
best_balanced_accuracy <- balanced_accuracies[best_lambda_index]
best_num_predictors <- num_predictors[best_lambda_index]

cat("Best Lambda:", best_lambda, "\n")
cat("Highest Balanced Accuracy:", best_balanced_accuracy, "\n")
cat("Number of Predictors for Best Lambda:", best_num_predictors, "\n")

```



### Forward stepwise selection (FSS) (???)

```{r, results='hold'}
fss_start_time <- Sys.time()

regfit_fwd <-  regsubsets(label ~ .,
                        data = task_two_df_first_523,
                        nvmax = 170,
                        method = "forward")
reg_summary <- summary(regfit_fwd)

fss_end_time <- Sys.time()
fss_runtime <- fss_end_time - fss_start_time
print(fss_runtime)

par(mfrow=c(2,2))
plot(reg_summary$rss, xlab="Number of Variables", ylab="RSS")

# Plot adjusted R2 vs Number of Variables
plot(reg_summary$adjr2, xlab="Number of Variables", ylab="Adjusted RSq", type="l")
# which_max() function is used to identify the location of the maximum point of a vector
best_model = which.max(reg_summary$adjr2)
# Plot a red dot to indicate the model with the largest adjusted R2
points(best_model, reg_summary$adjr2[best_model], col="red", cex=2, pch=20)

## In a similar fashion, we can plot Cp and BIC
plot(reg_summary$cp, xlab="Number of Variables", ylab="Cp", type="l")
best_model = which.min(reg_summary$cp)
points(best_model, reg_summary$cp[best_model], col="red", cex=2, pch=20)

plot(reg_summary$bic, xlab="Number of Variables", ylab="BIC", type="l")
best_model = which.min(reg_summary$bic)
points(best_model, reg_summary$bic[best_model], col="red", cex=2, pch=20)

```

```{r, fig.height = 20, fig.width = 50, fig.align = "center"}
par(mfrow = c(2, 2))
plot(regfit_fwd, scale = "r2")
plot(regfit_fwd, scale = "adjr2")
plot(regfit_fwd, scale = "Cp")
plot(regfit_fwd, scale = "bic")
```


### Random forest (???)

```{r}

```


## T2.3: additional feature selection methods

Train and evaluate up to three additional feature selection methods, 
other than those in T2.2, which may include methods not covered in the course, 
with the goal of achieving high balanced accuracy using a minimal number of 
selected features. Provide an explanation of each new method.

For T2.2 and T2.3, your evaluation results should include the balanced accuracy 
and the number of selected features achieved by each method for various numbers 
of selected features. You may also compare the balanced accuracy achieved 
by different methods for a similar number of selected features.

### Elastic net (not covered in class) (???)

```{r}

```


### [[Tree-Based models]] (i.e., either XGBoost or LightGBM) (???)

```{r}

```


### (Sparse?) SVM (???)

Is the non-sparse SVM a feature-selection method?

```{r}

```




